\Title

### 2013-08-17 synmut.tex
Fixes in the bulk, ready for me (Fabio)

\newpage
\section*{Abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$
T-cells or neutralizing antibodies target a new epitope, the virus often
escapes via nonsynonymous mutations that impair recognition. Synonymous
mutations do not affect this interplay and are often assumed to be
neutral. We test this assumption by tracking synonymous mutations in
longitudinal intrapatient data from the \shankaregion{} part of the
\env{} gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data from SHAPE assays, we find that
synonymous mutations that disrupt base pairs in RNA stems flanking the
variable loops of gp120 are more likely to be lost than other synonymous
changes: these RNA hairpins might be important for HIV-1.  Computational
modeling indicates that most synonymous mutations in this genomic region
have a selection coefficient of the order of $-0.003$, and that the
majority of synonymous mutations need to be slightly deleterious to be
consistent with the data. This weak selection against synonymous
substitutions does not result in a strong pattern of conservation in
cross sectional data, but slows down the rate of evolution
considerably. Our findings are consistent with the notion that large
scale patterns of RNA structure are functionally relevant, while the
precise base pairing pattern is not.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimates of time to fixation/extinction, SNVs were binned by
frequency and the time to first reaching either fixation or extinction was
stored. The fixation probability was determined as the long-time limit of the
resulting curves. Mutations that reached high frequency but neither fixed nor
were lost were classified as ``floating'', with one exception: if they first
reached high frequencies within 3 years of the last time point, it was assumed
they had not had sufficient time to settle, so they were discarded.

@@

Another crucial ingredient for our simulations is the fitness landscape,
i.e., the effects of mutations on fitness and possible interactions
between mutations. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $\alpha$ or a deleterious effect $s_d$ with probability
$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 1.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects of $-???$. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. The rate of escape in chronic infection is
reported to be on the order of a few percent
\citep{Asquith:2006p28003,moore_limited_2009}, consistent with the
finding that virus is neutralized by serum from 3-6 month later
\citep{richman_rapid_2003}. We simulated escape rates between $2 \cdot
10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
The rate at which new anti-body challenges arise is more difficult to
quantify, but a substantial fraction of non-synonymous substitutions in gp120 are
probably driven by escape. Since the sequence of the C2-V5 region of
gp120  changes at a rate of about 1\% per year
\citep{shankarappa_consistent_1999}, this rate should be on the order of
$k_A=0.005$ per day. We simulate a range from $10^{-4}$ to $10^{-2}$ per day.

For the models with competition within epitopes, a complex epistatic fitness
landscape was designed such that each single mutant is sufficient for full
escape. Specifically, each mutation has an additive effect equal to the
escape rate, but interacts with all other escape mutations in the same
epitope with a negative effect of the same magnitude. 
Higher order terms were included to make sure that not
only double mutants, but all k-mutants with $k \geq 1$ had the same fitness (see
supplementary materials). To model recognition of escape variants by the
evolving immune system, the beneficial effect of an escape mutation was set
to its previous cost of -???? with a probability per generation proportional to
the frequency of the escape variant.
TODO

\subsubsection*{Sampling and analysis}
To obtain reasonable sampling of a particular parameter combination we
ran simulation for 6000 days and we repeated each run 100 times with
different seeds for the random number generator. Both full-length HIV-1
genomes and \env{}-only simulations were performed and yielded
comparable results.  Populations were initialized with a homogeneous
founder population and were kept at an average size of $N=10^4$
throughout the simulation. After 30 generations of burn-in to create
genetic diversity, new epitopes were introduced at a constant rate
$k_A$.  The simulations were repeated 1000 times with a seven dimensional
Latin hypercube sampling scheme \citep{mckay_comparison_1979}.  For each
parameter, 1000 random values were chosen from a distribution uniform in
log-space and bounded by the ranges given for each of these parameters
above. These 1000 values for each parameter were combined using a Latin
hypercube sampling scheme.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation of some deleterious synonymous variants is not fully
unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
triangle line versus the green square line in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,wei_antibody_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
cohort.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution for many different combinations of
effects of synonymous mutations, $s_d$, and the fraction of synonymous
sites that are neutral, $\alpha$, escape rates and number of targeted epitopes. In
addition, we vary the mutation rate, recombination rate and population
size, see methods.  Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Figure \ref{fig:simsfig} shows
the distribution of deleterious effects $s_d$ and neutral fraction
$\alpha$ for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that in
order to be consistent with the data only
a minority of synonyous sites ($\lesssim 0.2$) can be neutral  and that
the deleterious effects of the remainder is on the order of $|s_d| \sim
0.003$.

@@

The above simulations show that hitchhiking with favoured nonsynonymous
variants can explain the observated pattern of deleterious synonymous SNVs that
rarely fix. While our model is already quite complicated with many
poorly known parameters that make interpretation of simulation results
challenging, we know that the actual evolutionary dynamics within an
infected person is much more complicated still. In our model, 
 escape mutations are unconditionally beneficial and almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the observed fixation probability of
nonsynonymous SNVs at high frequency often disappear again
(\FIG{fixp2}). One possible reason for this discrepancy is the fact that
the fitness landscape in our model is static: Once a site is designated
as targeted by the immune system, mutations at this site are beneficial
forever. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

@@

Furthermore, several different escape mutations might arise within the same epitope.
Their benefits are not additive, because each of
them is essentially sufficient to escape and no additional benefit is
gained from combining them. As a consequence, several escape SNVs can rise
to high frequency rapidly, while the one with the smallest cost in terms
of replication, packaging, etc.~is most likely to eventually fix, while
all others are lost. The emergence of multiple competing escape variants
in HIV-1 infections has been shown \citep{moore_limited_2009,
bar_early_2012,wei_antibody_2003}.  This scenario has been explicitly
observed in the evolution of resistance to 3TC, where the mutation M184V
is often preceded by M184I \citep{hedskog_dynamics_2010} or in the
context of CTL escape mutations where multiple variants rise to high
frequencies rapidly but most of the are lost again
\citep{fischer_transmission_2010}. Consequently, there could be many
non-synonymous mutations are only beneficial until the viral population
has found a ``better solution'' and subsequently disappear.

We implemented within epitope
competition in the model by allowing for multiple escape mutations per
epitope that do not provide additional benefit to the virus when
combined. Again, we found that the rapid rise of several non-synonymous
mutations to intermediate frequencies allows deleterious synonymous
mutations to hitch-hike, while the fixation probability of non-synonymous
mutations is reduced. With roughly six mutations per epitope,
the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios, time-dependent selection and
competition between equivalent escape pathways, are not exclusive and
possibly both important in HIV-1 evolution. Our modeling merely suggest
possible scenarios and to determine to what extend
either of the of these effects contribute to the observed pattern, more
data is required.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these SNVs have
deleterious effects on the order of $s_d = 0.003$ and that the majority
of all synonymous mutations is deleterious. In absense of hitch-hiking
in a large population, deleterious mutations with effect $s_d$ should be
at a frequency $\mu/s_d\approx 0.01$. The fact that we see many
synonymous mutation at high frequency that nevertheless disappears over
a time scale of a year suggests that they are brought to high frequency
by hitch-hiking on favorable genetic backgrounds. 
While both the rise and fall of synonymous SNV is subject to hichhiking,
the consistent loss of these variants conditional on being at high
frequency implies selection against these SNVs.
Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections. 
The fixation probabilities and the sojourn times of SNVs represent a
rich and simple summary statistics of longitudinal sequence data. Most
importantly, these statistics are informative even in absence of a
neutral control and thus appropriate to analyze properties of synonymous
sites.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn_latin_contour.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous SNVs. This parameter also reduces synonymous
diversity, measured by the probability of a SNV to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 3000 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$,
rate of introduction of new epitopes, population size, mutation rate, and
recombination rate (see methods). The density of simulations that reproduce
the synonymous diversity and fixation patterns observed in data are shown.
These simulations demonstrate that deleterious effects are around $-0.003$
and a large fraction of the synonymous mutations needs to be deleterious. This
picture is also confirmed by the analysis of the marginal distributions of the
single parameters (see \figurename~S\marginals).}
\label{fig:simheat}
\end{center}
\end{figure}

>>>>

\newpage
\section*{Abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$
T-cells or neutralizing antibodies target a new epitope, the virus often
escapes via nonsynonymous mutations that impair recognition. Synonymous
mutations do not affect this interplay and are often assumed to be
neutral. We test this assumption by tracking synonymous mutations in
longitudinal intrapatient data from the \shankaregion{} part of the
\env{} gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data from SHAPE assays, we find that
synonymous mutations that disrupt base pairs in RNA stems flanking the
variable loops of gp120 are more likely to be lost than other synonymous
changes: these RNA hairpins might be important for HIV-1.  Computational
modeling indicates that, to be consistent with the data, a large fraction
of synonymous mutations in this genomic region needs to be deleterious,
with a selection coefficient of the order of $-0.002$. This weak selection against synonymous
substitutions does not result in a strong pattern of conservation in
cross sectional data, but slows down the rate of evolution
considerably. Our findings are consistent with the notion that large
scale patterns of RNA structure are functionally relevant, while the
precise base pairing pattern is not.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimates of time to fixation/extinction, single nucleotide variants
(SNVs) were binned by
frequency and the time to first reaching either fixation or extinction was
stored. The fixation probability was determined as the long-time limit of the
resulting curves. Mutations that reached high frequency but neither fixed nor
were lost were classified as ``floating'', with one exception: if they first
reached high frequencies within 3 years of the last time point, it was assumed
they had not had sufficient time to settle, so they were discarded.

@@

Another crucial ingredient for our simulations is the fitness landscape,
i.e., the effects of mutations on fitness and possible interactions
between mutations. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $\alpha$ or a deleterious effect $s_d$ with probability
$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 1.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects of $-0.01$ (simulations with a larger cost of $-0.1$ were also
performed and produced similar results). At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. The rate of escape in chronic infection is
reported to be on the order of a few percent
\citep{Asquith:2006p28003,moore_limited_2009}, consistent with the
finding that virus is neutralized by serum from 3-6 month later
\citep{richman_rapid_2003}. We simulated escape rates between $2 \cdot
10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
The rate at which new antibody challenges arise is more difficult to
quantify, but a substantial fraction of nonsynonymous substitutions in gp120 are
probably driven by escape. Since the sequence of the C2-V5 region of
gp120  changes at a rate of about 1\% per year
\citep{shankarappa_consistent_1999}, this rate should be on the order of
$k_A=0.005$ per day. We simulate a range from $10^{-4}$ to $10^{-2}$ per day.

For the models with competition within epitopes, a complex epistatic fitness
landscape was designed such that each single mutant is sufficient for full
escape. Specifically, each mutation has an additive effect equal to the
escape rate, but interacts with all other escape mutations in the same
epitope with a negative effect of the same magnitude. 
Higher order terms were included to make sure that not
only double mutants, but all k-mutants with $k \geq 1$ had the same fitness (see
supplementary materials). To model recognition of escape variants by the
evolving immune system, the beneficial effect of an escape mutation was set
to its previous cost of $-0.01$ with a probability per generation proportional to
the frequency of the escape variant.

\subsubsection*{Sampling and analysis}
To obtain reasonable sampling of a particular parameter combination we
ran simulation for 6000 days and we repeated each run 100 times with
different seeds for the random number generator. Both full-length HIV-1
genomes and \env{}-only simulations were performed and yielded
comparable results. Populations were initialized with a homogeneous
founder population and were kept at an average size of $N=10^4$
throughout the simulation. After 30 generations of burn-in to create
genetic diversity, new epitopes were introduced at later times according 
to an exponential distribution with rate
$k_A$. The simulations were repeated 3000 times with a seven dimensional
Latin hypercube sampling scheme \citep{mckay_comparison_1979}, bounded by the ranges
given for each of these parameters above.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation of some deleterious synonymous variants is not fully
unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
and green lines in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,wei_antibody_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
cohort.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution for many different combinations of
effects of synonymous mutations, $s_d$, and the fraction of synonymous
sites that are neutral, $\alpha$, escape rates and number of targeted epitopes. In
addition, we vary the mutation rate, recombination rate and population
size (see methods).  Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Figure \ref{fig:simsfig} shows
the distribution of deleterious effects $s_d$ and neutral fraction
$\alpha$ for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that in
order to be consistent with the data only
a minority of synonyous sites ($\lesssim 0.2$) can be neutral  and that
the deleterious effects of the remainder is on the order of $|s_d| \sim
0.002$.

@@

The above simulations show that hitchhiking on favoured nonsynonymous
variants can recapitulate the observated pattern of deleterious synonymous SNVs that
rarely fix. While our model is already quite complicated with many
poorly known parameters that make interpretation of simulation results
challenging, we know that the actual evolutionary dynamics within an
infected person is much more complicated still. In our model, 
 escape mutations are unconditionally beneficial and almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the observed fixation probability of
nonsynonymous SNVs at high frequency often disappear again
(\FIG{fixp2}). One possible reason for this discrepancy is the fact that
the fitness landscape in our model is static: Once a site is designated
as targeted by the immune system, mutations at this site are beneficial
forever. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

@@

Furthermore, several different escape mutations might arise within the same epitope.
Their benefits are not additive, because each of
them is essentially sufficient to escape and no additional benefit is
gained from combining them. As a consequence, several escape SNVs can rise
to high frequency rapidly, while the one with the smallest cost in terms
of replication, packaging, etc.~is most likely to eventually fix, while
all others are lost. The emergence of multiple competing escape variants
in HIV-1 infections has been shown \citep{moore_limited_2009,
bar_early_2012,wei_antibody_2003}.  This scenario has been explicitly
observed in the evolution of resistance to 3TC, where the mutation M184V
is often preceded by M184I \citep{hedskog_dynamics_2010} or in the
context of CTL escape mutations where multiple variants rise to high
frequencies rapidly but most of the are lost again
\citep{fischer_transmission_2010}. Consequently, there could be many
nonsynonymous mutations are only beneficial until the viral population
has found a ``better solution'' and subsequently disappear.

We implemented within epitope
competition in the model by allowing for multiple escape mutations per
epitope that do not provide additional benefit to the virus when
combined. Again, we found that the rapid rise of several nonsynonymous
mutations to intermediate frequencies allows deleterious synonymous
mutations to hitchhike, while the fixation probability of nonsynonymous
mutations is reduced. With roughly six mutations per epitope,
the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios, time-dependent selection and
competition between equivalent escape pathways, are not exclusive and
possibly both important in HIV-1 evolution. Our modeling merely suggest
possible scenarios and to determine to what extend
either of the of these effects contribute to the observed pattern, more
data is required.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these SNVs have
deleterious effects on the order of $s_d = 0.002$ and that the majority
of all synonymous mutations is deleterious. In absense of hitchhiking
in a large population, deleterious mutations with effect $s_d$ should be
at a frequency $\mu/s_d\approx 0.01$. The fact that we see many
synonymous mutation at high frequency that nevertheless disappears over
a time scale of a year suggests that they are brought to high frequency
by hitchhiking on favorable genetic backgrounds. 
While both the rise and fall of synonymous SNV is subject to hichhiking,
the consistent loss of these variants conditional on being at high
frequency implies selection against these SNVs.
Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{in vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections. 
The fixation probabilities and the sojourn times of SNVs represent a
rich and simple summary statistics of longitudinal sequence data. Most
importantly, these statistics are informative even in absence of a
neutral control and thus appropriate to analyze properties of synonymous
sites.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn_latin_contour.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous SNVs. This parameter also reduces synonymous
diversity, measured by the probability of a SNV to be found at
intermediate frequencies $P_\text{interm}$ (first inset). The area observed in
the data is also shown (second inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 3000 simulations with random combinations of parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$,
rate of introduction of new epitopes, population size, mutation rate, and
recombination rate (see methods). The density of simulations that reproduce
the synonymous diversity and fixation patterns observed in data are shown.
These simulations demonstrate that deleterious effects are around $-0.002$
and a large fraction of the synonymous mutations needs to be deleterious. This
picture is also confirmed by the analysis of the marginal distributions of the
single parameters (see \figurename~S\marginals).}
\label{fig:simheat}
\end{center}
\end{figure}
